\huge I don't know, are you sure we want to do this? Sea level adaptation decisions under uncertainty

### 2017-01-06 techreport/paper.tex
Technical Report

%% To submit your paper:
\documentclass[draft,linenumbers]{agujournal}
\draftfalse

%% For final version.
% \documentclass{agujournal}

\journalname{Water Resources Research}

\usepackage{url, bbm, comment}

@@

\title{I don't know, are you sure we want to do this?\\
Sea level adaptation decisions under uncertainty}

\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, P. Skougaard Kaspersen\affil{2} and K. de Bruin\affil{3,4}}

\affiliation{1}{Norwegian Computing Center, Oslo, Norway}
\affiliation{2}{Technical University of Denmark, Copenhagen, Denmark}
\affiliation{3}{Center for International Climate and Environmental Research, Oslo, Norway}
\affiliation{4}{Wageningen Environmental Research, Wageningen, The Netherlands}

\correspondingauthor{T. L. Thorarinsdottir}{thordis@nr.no}

\begin{keypoints}
\item Decisions on adaptation measures need to take careful account of all uncertainties
\item Modeling local sea level rise is essential
\item Decisions under uncertainty do not have to be be based on precise costs
\end{keypoints}

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action. Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations. Different components of uncertainty are visualized. We demonstrate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller. 
\end{abstract}

%% ------------------------------------------------------------------------ %%
%  BEGIN ARTICLE
%% ------------------------------------------------------------------------ %%

@@

We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies. 
We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
\begin{linenomath*}
\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
\end{linenomath*}
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
\begin{linenomath*}
\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
\end{linenomath*}
where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors.

@@

We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
\begin{linenomath*}
  \begin{equation}\label{eq:Burr}
  f_{\textup{d}, t_0} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
  \end{equation}
\end{linenomath*}
for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter. The Burr distribution has a heavy upper tail and is commonly used to model damage loss, see e.g. \cite{Klugman&2012}. The parameters of the distribution are estimated using historical data for annual storm surge damage. Data prior to 2015 are adjusted to the 2015 level using the consumer price index. After adjustment, we assume stationarity over the period and independence between years.

Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d. values from the estimated distribution $\hat{F}_{\textup{d}, t_0}$.  By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l}) } \leq x \Bigg\}
  \]
\end{linenomath*}
for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
\begin{linenomath*}
  \[
  d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})}
  \]
\end{linenomath*}
and similarly for the cumulative damage.

@@

We assume that changes in sea level have a multiplicative effect on the annual damage cost. That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
\begin{linenomath*}
  \[
  g(s_{t_i} | \boldsymbol \beta ) d_{t_i},
  \]
\end{linenomath*}
where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$. \cite{Hallegatte&2013} estimate a similar effect function valid in 2050 for $s \in \{0,20,40\}$ for 136 coastal cities. Here, we use their results for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseilles, Naples, Porto, Rotterdam and Stockholm.  To obtain a city-specific effect function for a large range of sea level anomalies we employ a linear extrapolation as shown in Figure~\ref{fig:EffectFct}. We then obtain a sample of effect functions $\{g(\cdot|\boldsymbol \beta^{(j)})\}_{j=1}^J$ by sampling with replacement from this ensemble of trajectories with all 15 ensemble members considered equally probable.

@@

Let further $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ denote a sample of projections for annual sea level anomalies compared to the 2015 value. An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
\begin{linenomath*}
  \begin{equation}\label{eq:FutureDamage}
    \hat{F}_{\textup{d}, t_i}^\textup{s} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}.
  \end{equation}
\end{linenomath*}

The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures. In addition, we can incorporate an adaptation measure of cost $C$ that protects against $K$ cm of increased sea level from year $t_k$ onward. This results in a damage distribution given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i}^{\textup{s}, \textup{a}_k} (x) =
    \begin{cases}
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i < t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} + C}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i = t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i > t_k. \\
    \end{cases}
  \]
\end{linenomath*}

@@

%  ACKNOWLEDGMENTS
\begin{acknowledgments}
This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL). The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
\end{acknowledgments}

>>>>

\documentclass[note,screen,british,11pt]{nrdoc}

\usepackage{url, bbm, comment, verbatim}

% -------------------------------------------------
% BEGIN nrdoc special commands
\ifx\nrdocument\undefined\relax\else
\reportnumber{samba/02/17}
\keywords{Sea level rise; Adaptation; Decision making}
\projectnumber{220730}
\project{eSACP}
\availability{Open}
\frontpagefigure{sealevel-tim-marshall} 
\target{Decision makers and scientists}
\researchfield{Statistics, Climatology, Environmental Economics}
\fi
% END nrdoc special commands
% ------------------------------------------------

\makeindex

@@

\title{\huge I don't know, are you sure we want to do this? Sea level adaptation decisions under uncertainty}

\nrtitle{I don't know, are you sure we want to do this? Sea level adaptation decisions under uncertainty}

\shorttitle{Sea level adaptation decisions under uncertainty}

\author{Thordis L. Thorarinsdottir \email{thordis@nr.no} \and Peter Guttorp \email{peter.guttorp@nr.no} \and Martin Drews \email{mard@dtu.dk} \and Per Skougaard Kaspersen \email{pskk@dtu.dk} \and Karianne de Bruin \email{karianne.debruin@cicero.oslo.no}}

\aboutauthors{Thordis L. Thorarinsdottir is Chief Research Scientist and Peter Guttorp is Professor II at the Norwegian Computing Center in Oslo, Norway. Martin Drews is a Senior Scientist and Per Skougaard Kaspersen is Postdoc at DTU Management Engineering, Technical University of Denmark in Copenhagen, Denmark. Karianne de Bruin is a Senior Researcher at the Center for International Climate and Environmental Research in Oslo, Norway and Researcher at Wageningen Environmental Research in Wageningen, The Netherlands}

\date{\today}

\maketitle

\begin{abstract}
  Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action. Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations. Different components of uncertainty are visualized. We demonstrate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller.

  Front page photo is by Tim Marshall (\url{https://unsplash.com/@timmarshall}). 
\end{abstract}

@@

We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies. 
We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors.

@@

We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
  \begin{equation}\label{eq:Burr}
  f_{\textup{d}, t_0} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
  \end{equation}
for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter. The Burr distribution has a heavy upper tail and is commonly used to model damage loss, see e.g. \cite{Klugman&2012}. The parameters of the distribution are estimated using historical data for annual storm surge damage. Data prior to 2015 are adjusted to the 2015 level using the consumer price index. After adjustment, we assume stationarity over the period and independence between years.

Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d. values from the estimated distribution $\hat{F}_{\textup{d}, t_0}$.  By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
  \[
  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l}) } \leq x \Bigg\}
  \]
for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
  \[
  d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})}
  \]
and similarly for the cumulative damage.

@@

We assume that changes in sea level have a multiplicative effect on the annual damage cost. That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
  \[
  g(s_{t_i} | \boldsymbol \beta ) d_{t_i},
  \]
where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$. \cite{Hallegatte&2013} estimate a similar effect function valid in 2050 for $s \in \{0,20,40\}$ for 136 coastal cities. Here, we use their results for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseilles, Naples, Porto, Rotterdam and Stockholm.  To obtain a city-specific effect function for a large range of sea level anomalies we employ a linear extrapolation as shown in Figure~\ref{fig:EffectFct}. We then obtain a sample of effect functions $\{g(\cdot|\boldsymbol \beta^{(j)})\}_{j=1}^J$ by sampling with replacement from this ensemble of trajectories with all 15 ensemble members considered equally probable.

@@

Let further $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ denote a sample of projections for annual sea level anomalies compared to the 2015 value. An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
  \begin{equation}\label{eq:FutureDamage}
    \hat{F}_{\textup{d}, t_i}^\textup{s} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}.
  \end{equation}

The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures. In addition, we can incorporate an adaptation measure of cost $C$ that protects against $K$ cm of increased sea level from year $t_k$ onward. This results in a damage distribution given by
  \[
  \hat{F}_{\textup{d}, t_i}^{\textup{s}, \textup{a}_k} (x) =
    \begin{cases}
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i < t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} + C}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i = t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i > t_k. \\
    \end{cases}
  \]

@@

%  ACKNOWLEDGMENTS
\section*{Acknowledgments}
This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL). The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
